Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-20 paper.tex
reduction rules for S(C)
remove label of succ/fail
filled in missing rules in S(C)
highlight S(C) rules
minor fix
minor change

@@

\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}

@@

@@

@@

The dynamic semantics of a cast calculus is typically defined with a
reduction semantics. Here we use a CEK machine
\citep{felleisen1986control} instead because the first author is more
familiar with CEK machines and believes that a CEK machine is more
convenient to use for the space-efficient semantics in
Section~\ref{sec:framework:cek} (space-efficiency is about compressing
continuations). So using an abstract machine for the cast calculi in
this section makes it easier prove correctness of the space-efficient
machines. Of course, one should prove that the abstract machine
presented here is equivalent to the standard reduction semantics for
cast calculi, but we have not yet done so.

@@

\redrule{
\sOOreturn{v_2}{(\kOOappII{(\vOOfun{\rho}{x}{e})}{\kappa})}}{
\sOOinspect{e}{\rho[x:=v_2]}{\kappa}}{}
	\hiredrule{
		\sOOreturn{v_1}{(\mathtt{app_2} \; \vOOcast{v_2}{
				\cOOcast{\POOfun{T_1}{T_2}}{l}{\POOfun{T_3}{T_4}}
			} \; \kappa)}
	}{
		\sOOreturn{v_1}{
			\langle\cOOcast{T_3}{l}{T_1}\rangle
			(\mathtt{app_2} \; v_2 \; 
			\langle\cOOcast{T_2}{l}{T_4}\rangle \kappa)}
	}{}
	\hiredrule{
		\sOOreturn{
			\vOOcast{v}{\cOOcast{\POOprod{T_1}{T_2}}{l}{
					\POOprod{T_3}{T_4}}}
		}{(\mathtt{fst} \; \kappa)}
	}{
		\sOOreturn{v}{(
			\mathtt{fst} \;
			\langle \cOOcast{T_1}{l}{T_3} \rangle \kappa
			)}
	}{}

@@

	\redrule{
		\highlight{\sOOreturn{v}{(
			\mathtt{cast} \; c \; \kappa
			)}}
	}{
\highlight{          
\begin{cases}
	\sOOreturn{u}{\kappa} & \sidecond{applyCast(v,c) = \rOOsucc{u}}
	\\
	\sOOhalt{(\oOOblame{l})} & \sidecond{applyCast(v,c) = \rOOfail{l}}
\end{cases}}
	}{}
\redrule{
\sOOreturn{v}{\kOOmt}}{
\sOOhalt{observe(v)}}{}
	\end{array}
	\]

@@

\todo[inline]{There are missing transition rules for fst and snd. -JS}

@@

	\stxrule{v}{values}{
		\hcvOOtt \mid
		\hcvOOfun{c}{\rho}{x}{e}{c} \mid
		\hcvOOcons{v}{c}{v}{c}
	}
	\stxrulecont{
		\hcvOOinl{v}{c} \mid
		\hcvOOinr{v}{c} \mid
		\hcvOOinj{P}{v}
	}
	\stxrule{r}{cast results}{
		\rOOsucc{v} \mid
		\rOOfail{l}
	}
	\stxrule{s}{states}{
		\sOOinspect{e}{\rho}{\kappa} \mid{}
		\sOOreturn{v}{\kappa} \mid{}
		\sOOhalt{o}
	}
	\stxrule{\kappa}{continuation}{
		\langle c \rangle k
	}
	\stxrule{k}{pre-continuations}{
		\hckOOmt \mid{}
		\mathtt{cons_1} \; e \; \rho \; \kappa \mid{}
		\mathtt{cons_2} \; v \; \kappa \mid{}
		\mathtt{inl} \; \kappa \mid{}
		\mathtt{inr} \; \kappa
	}
	\stxrulecont{
		\mathtt{app_1} \; e \; \rho \; \kappa \mid{}
		\mathtt{app_2} \; v \; \kappa \mid{}
		\mathtt{fst} \; \kappa \mid{}
		\mathtt{snd} \; \kappa
	}
	\stxrulecont{
		\mathtt{case_1} \; e \; e \; \rho \; \kappa \mid
		\mathtt{case_2} \; v \; e \; \rho \; \kappa \mid{}
		\mathtt{case_3} \; v \; v \; \rho \; \kappa
	}
	\end{array}
	\]

@@

	Reduction \fbox{$ \judgeSreduce{C}{s}{s} $}
	\[
	\begin{array}{rclr}
	& \vdots \\
	\redruleS{
		\sOOinspect{(\eOOlam{T_1}{T_2}{x}{e})}{\rho}{\kappa}
	}{
		\sOOreturn{(\hcvOOfun{id(T_1)}{\rho}{x}{e}{id(T_2)})}{\kappa}
	}{}
	\redruleS{
		\sOOinspect{\eOOcast{e}{T_1}{l}{T_2}}{\rho}{\kappa}
	}{
		\sOOinspect{e}{\rho}{ext(cast(T_1,l,T_2),\kappa)}
	}{}
	\redruleS{
		\sOOinspect{(\eOOcons{e_1}{e_2})}{\rho}{\kappa}
	}{
		\sOOinspect{e_1}{\rho}{cont(\hckOOconsI{e_2}{\rho}{\kappa})}
	}{}
	\redruleS{
		\sOOreturn{(\hcvOOfun{c_1}{\rho}{x}{e}{c_2})}{(c,\hckOOappII{v}{\kappa})}
	}{
		\begin{cases}
		\sOOinspect{e}{\rho[x:=v']}{ext(c_2,\kappa)} \\
		\;\;\;\;\;\;\;\;\;\sidecond{applyCast(v,c_1) = \rOOsucc{v'}} 
		\\
		\sOOhalt{(\oOOblame{l})} \\
		\;\;\;\;\;\;\;\;\;\sidecond{applyCast(v,c_1) = \rOOfail{l}}
		\end{cases}

	}{}
	\end{array}
	\]

@@

	Evaluation \fbox{$ \judgeSeval{C}{e}{o} $}
	\[
	\inference{
		\judgeSreduceTrans{C}{
			\sOOinspect{e}{\emptyset}{cont(\hckOOmt)}
		}{
			\sOOhalt{o}
		}		
	}{
		\judgeSeval{C}{e}{o}
	}
	\]

@@

\todo[inline]{KC needs to add missing rules in and describe them. -KC}
We list a fraction of reduction rules due to space limitation.
When values are constructed, their hypercoercion parts are filled with outputs 
of $ id $. For instance, when a function value is constructed, its first part 
and last part are initialized to $ id(T_1) $ and $ id(T_2) $ respectively.
When evaluating a cast expression, the current continuation is extended with a 
hypercoercion constructed by $ cast $. $ ext $ composes the new hypercoercion 
with the hypercoercion on the top of the continuation.
When evaluating a compound expression, the machine firstly construct the new 
pre-continuation, then turn it to a continuation by adding an identity 
hypercoercion at the top. For instance, when evaluating a \texttt{cons} 
expression, the machine firstly construct $ \hckOOconsI{e_2}{\rho}{\kappa} $, 
the new pre-continuation, then call $ cont $, which adds an identity cast to 
form a continuation. 
When a function call happens, the machine firstly cast the operand. If the 
casting succeeds, the machine then evaluate the body in the extended 
environment and the extended continuation. If the casting fails, the machine 
then halts with the blame label.

>>>>

\newcommand{\hifunrule}[3]{\highlight{#1} &\highlight{=}& \highlight{#2} & 
\highlight{#3}\\}

@@

\newcommand{\rOOsucc}[1]{#1}
\newcommand{\rOOfail}[1]{#1}
%\newcommand{\rOOsucc}[1]{\mathtt{succ}\;#1}
%\newcommand{\rOOfail}[1]{\mathtt{fail}\;#1}

@@

\newcommand{\hiredruleS}[3]{\highlight{#1} & 
\highlight{\longmapsto_{\mathcal{S}(C)}} & \highlight{#2} & \highlight{#3} \\}

@@

\newcommand{\continue}[2]{continue(#2,#1)}

@@

The dynamic semantics of a cast calculus is typically defined with a
reduction semantics. Here we use a CEK machine
\citep{felleisen1986control} instead because the first author is more
familiar with CEK machines and believes that a CEK machine is more
convenient to use for the space-efficient semantics in
Section~\ref{sec:framework:cek} (a major point of space-efficiency is about 
compressing continuations). So using an abstract machine for the cast calculi in
this section makes it easier prove correctness of the space-efficient
machines. Of course, one should prove that the abstract machine
presented here is equivalent to the standard reduction semantics for
cast calculi, but we have not yet done so.

@@

\redrule{
\sOOreturn{v_2}{(\kOOappII{(\vOOfun{\rho}{x}{e})}{\kappa})}}{
\sOOinspect{e}{\rho[x:=v_2]}{\kappa}}{}
	\hiredrule{
		\sOOreturn{v_1}{(\mathtt{app_2} \; \vOOcast{v_2}{
				\cOOcast{\POOfun{T_1}{T_2}}{l}{\POOfun{T_3}{T_4}}
			} \; \kappa)}
	}{
		\sOOreturn{v_1}{
			\langle\cOOcast{T_3}{l}{T_1}\rangle
			(\mathtt{app_2} \; v_2 \; 
			\langle\cOOcast{T_2}{l}{T_4}\rangle \kappa)}
	}{}
	\redrule{
	\sOOreturn{
		(\vOOcons{v_1}{v_2})
	}{(\mathtt{fst} \; \kappa)}
	}{
	\sOOreturn{v_1}{\kappa}
	}{}
	\hiredrule{
		\sOOreturn{
			\vOOcast{v}{\cOOcast{\POOprod{T_1}{T_2}}{l}{
					\POOprod{T_3}{T_4}}}
		}{(\mathtt{fst} \; \kappa)}
	}{
		\sOOreturn{v}{(
			\mathtt{fst} \;
			\langle \cOOcast{T_1}{l}{T_3} \rangle \kappa
			)}
	}{}
	\redrule{
	\sOOreturn{
		(\vOOcons{v_1}{v_2})
	}{(\mathtt{snd} \; \kappa)}
	}{
	\sOOreturn{v_2}{\kappa}
	}{}

@@

	\redrule{
		\highlight{\sOOreturn{v}{(
			\mathtt{cast} \; c \; \kappa
			)}}
	}{
\highlight{          
\begin{cases}
	\sOOreturn{v'}{\kappa} & \sidecond{applyCast(v,c) = \rOOsucc{v'}}
	\\
	\sOOhalt{(\oOOblame{l})} & \sidecond{applyCast(v,c) = \rOOfail{l}}
\end{cases}}
	}{}
\redrule{
\sOOreturn{v}{\kOOmt}}{
\sOOhalt{observe(v)}}{}
	\end{array}
	\]

@@

\todo[inline]{There are missing transition rules for fst and snd. -JS
\\---\\
Fixed. -KC}

@@

	\stxrule{v}{values}{
		\hcvOOtt \mid
		\hcvOOfun{c}{\rho}{x}{e}{c} \mid
		\hcvOOcons{v}{c}{v}{c}
	}
	\stxrulecont{
		\hcvOOinl{v}{c} \mid
		\hcvOOinr{v}{c} \mid
		\hcvOOinj{P}{v}
	}
%	\stxrule{r}{cast results}{
%		\rOOsucc{v} \mid
%		\rOOfail{l}
%	}
%	\stxrule{s}{states}{
%		\sOOinspect{e}{\rho}{\kappa} \mid{}
%		\sOOreturn{v}{\kappa} \mid{}
%		\sOOhalt{o}
%	}
	\stxrule{\kappa}{continuation}{
		\langle c \rangle k
	}
	\stxrule{k}{pre-continuations}{
		\hckOOmt \mid{}
		\mathtt{cons_1} \; e \; \rho \; \kappa \mid{}
		\mathtt{cons_2} \; v \; \kappa \mid{}
		\mathtt{inl} \; \kappa \mid{}
		\mathtt{inr} \; \kappa
	}
	\stxrulecont{
		\mathtt{app_1} \; e \; \rho \; \kappa \mid{}
		\mathtt{app_2} \; v \; \kappa \mid{}
		\mathtt{fst} \; \kappa \mid{}
		\mathtt{snd} \; \kappa
	}
	\stxrulecont{
		\mathtt{case_1} \; e \; e \; \rho \; \kappa \mid
		\mathtt{case_2} \; v \; e \; \rho \; \kappa \mid{}
		\mathtt{case_3} \; v \; v \; \rho \; \kappa
	}
	\end{array}
	\]

@@

	Reduction \fbox{$ \judgeSreduce{C}{s}{s} $}
	\[
	\begin{array}{rclr}
	\hiredruleS{
		\sOOinspect{\eOOcast{e}{T_1}{l}{T_2}}{\rho}{\kappa}
	}{
		\sOOinspect{e}{\rho}{ext(cast(T_1,l,T_2),\kappa)}
	}{}
	\redruleS{
		\sOOinspect{\eOOvar{x}}{\rho}{\kappa}
	}{	
		\sOOreturn{\rho[x]}{\kappa}
	}{}
	\redruleS{
		\sOOinspect{\eOOsole}{\rho}{\kappa}
	}{
		\sOOreturn{\hcvOOtt}{\kappa}
	}{}
	\hiredruleS{
		\sOOinspect{\eOOlam{T_1}{T_2}{x}{e}}{\rho}{\kappa}
	}{
		\sOOreturn{(\hcvOOfun{id(T_1)}{\rho}{x}{e}{id(T_2)})}{\kappa}
	}{}
	\redruleS{
		\sOOinspect{(\eOOcons{e_1}{e_2})}{\rho}{\kappa}
	}{
		\sOOinspect{e_1}{E}{}
	}{}
	\redruleS{
		\sOOinspect{(\eOOinl{e})}{\rho}{\kappa}
	}{
		\sOOinspect{e}{\rho}{cont(\kOOinl{\kappa})}
	}{}
	\redruleS{
		\sOOinspect{(\eOOinr{e})}{\rho}{\kappa}
	}{
		\sOOinspect{e}{\rho}{cont(\kOOinr{\kappa})}
	}{}
	\redruleS{
		\sOOinspect{(\eOOapp{e_1}{e_2})}{\rho}{\kappa}
	}{
		\sOOinspect{e_1}{\rho}{cont(\kOOappI{E}{e_2}{\kappa})}}{}

	\redruleS{
		\sOOinspect{(\eOOcar{e})}{\rho}{\kappa}}{
		\sOOinspect{e}{\rho}{cont(\kOOcar{\kappa})}}{}

	\redruleS{
		\sOOinspect{(\eOOcdr{e})}{\rho}{\kappa}}{
		\sOOinspect{e}{\rho}{cont(\kOOcdr{\kappa})}}{}

	\redruleS{
		\sOOinspect{(\eOOcase{e_1}{e_2}{e_3})}{\rho}{\kappa}}{
		\sOOinspect{e_1}{\rho}{cont(\kOOcaseI{E}{e_2}{e_3}{\kappa})}}{}
	\hiredruleS{
		\sOOreturn{v}{\langle c \rangle k}
	}{
	\begin{cases}
	continue(k,v') & \sidecond{applyCast(v,c) = \rOOsucc{v'}} 
	\\
	\sOOhalt{(\oOOblame{l})} & \sidecond{applyCast(v,c) = \rOOfail{l}}
	\end{cases}
	}{}
%\redruleS{
%	\sOOreturn{v_1}{(\kOOconsI{e_2}{\rho}{\kappa})}}{
%	\sOOinspect{e_2}{\rho}{cont(\kOOconsII{v_1}{\kappa})}}{}
%
%\redrule{
%\sOOreturn{v_2}{(\kOOconsII{v_1}{\kappa})}}{
%\sOOreturn{(\hcvOOcons{v_1}{id(T_1)}{v_2}{id(T_2)})}{\kappa}}{
%\\&&
%\sidecond{v_1 : T_1 \wedge v_2 : T_2}}
%
%\redrule{
%\sOOreturn{v}{(\kOOinl{\kappa})}}{
%\sOOreturn{(\hcvOOinl{v}{id(T)})}{\kappa}}{
%\\&&
%\sidecond{v : T}}
%
%\redrule{
%\sOOreturn{v}{(\kOOinr{\kappa})}}{
%\sOOreturn{(\hcvOOinr{v}{id(T)})}{\kappa}}{
%\\&&
%\sidecond{v : T}}
%
%\redrule{
%\sOOreturn{v_1}{(\kOOappI{e_2}{\rho}{\kappa})}}{
%\sOOinspect{e_2}{\rho}{cont(\kOOappII{v_1}{\kappa})}}{}
%
%	
%	& \vdots \\
%	\redruleS{
%		\sOOinspect{(\eOOlam{T_1}{T_2}{x}{e})}{\rho}{\kappa}
%	}{
%		\sOOreturn{(\hcvOOfun{id(T_1)}{\rho}{x}{e}{id(T_2)})}{\kappa}
%	}{}
%	\redruleS{
%		\sOOinspect{\eOOcast{e}{T_1}{l}{T_2}}{\rho}{\kappa}
%	}{
%		\sOOinspect{e}{\rho}{ext(cast(T_1,l,T_2),\kappa)}
%	}{}
%	\redruleS{
%		\sOOinspect{(\eOOcons{e_1}{e_2})}{\rho}{\kappa}
%	}{
%		\sOOinspect{e_1}{\rho}{cont(\hckOOconsI{e_2}{\rho}{\kappa})}
%	}{}
%	\redruleS{
%		
%\sOOreturn{(\hcvOOfun{c_1}{\rho}{x}{e}{c_2})}{(c,\hckOOappII{v}{\kappa})}
%	}{
%		\begin{cases}
%		\sOOinspect{e}{\rho[x:=v']}{ext(c_2,\kappa)} \\
%		\;\;\;\;\;\;\;\;\;\sidecond{applyCast(v,c_1) = \rOOsucc{v'}} 
%		\\
%		\sOOhalt{(\oOOblame{l})} \\
%		\;\;\;\;\;\;\;\;\;\sidecond{applyCast(v,c_1) = \rOOfail{l}}
%		\end{cases}
%		
%	}{}
	\end{array}
	\]

	Apply continuation
	\fbox{$ continue(k,v) = s $}
	\[
	\begin{array}{rclr}

\funrule{
	\continue{v_1}{(\kOOconsI{e_2}{\rho}{\kappa})}}{
	\sOOinspect{e_2}{\rho}{cont(\kOOconsII{v_1}{\kappa})}}{}

\hifunrule{
	\continue{v_2}{(\kOOconsII{v_1}{\kappa})}}{
	\sOOreturn{(\hcvOOcons{v_1}{id(T_1)}{v_2}{id(T_2)})}{\kappa}}{}

\hifunrule{
	\continue{v}{(\kOOinl{\kappa})}}{
	\sOOreturn{(\hcvOOinl{v}{id(T)})}{\kappa}}{}

\hifunrule{
	\continue{v}{(\kOOinr{\kappa})}}{
	\sOOreturn{(\hcvOOinr{v}{id(T)})}{\kappa}}{}

\funrule{
	\continue{v_1}{(\kOOappI{e_2}{\rho}{\kappa})}}{
	\sOOinspect{e_2}{\rho}{(\kOOappII{v_1}{\kappa})}}{}

%\funrule{
%	\continue{v_2}{(\kOOappII{(\hcvOOfun{c_1}{\rho}{x}{e}{c_2})}{\kappa})}}{
%applyCast(v_2,c_1) >>= \lambda v. \sOOinspect{e}{\rho[x:=v]}{ext(c,\kappa)}
%}{}

\hifunrule{
	\continue{v_2}{(\kOOappII{(\hcvOOfun{c_1}{\rho}{x}{e}{c_2})}{\kappa})}}{
\begin{cases}
	\sOOinspect{e}{\rho[x:=v']}{ext(c_2,\kappa)} \\
\;\;\;\;\;\sidecond{applyCast(v_2,c_1) = \rOOsucc{v'}}
\\
\sOOhalt{l} \\
\;\;\;\;\;\sidecond{applyCast(v_2,c_1) = \rOOfail{l}}
\end{cases}
}{}
\hifunrule{
	\continue{v_3}{(\mathtt{case_3}\;(\hcvOOinl{v}{c})\;
		(\hcvOOfun{c_1}{\rho}{x}{e}{c_2})
		\;\kappa)}
}{
	\sOOreturn{v}{cont(\kOOappII{(\hcvOOfun{seq(c,c_1)}{\rho}{x}{e}{c_2})}{\kappa})}
}{}

\hifunrule{
	\continue{(\hcvOOfun{c_1}{\rho}{x}{e}{c_2})}
	{(\mathtt{case_3}\;(\hcvOOinr{v}{c})\;v_2\;\kappa)}
}{
	\sOOreturn{v}{cont(\kOOappII{(\hcvOOfun{seq(c,c_1)}{\rho}{x}{e}{c_2})}{\kappa})}
}{}

\funrule{
	\continue{v}{\kOOmt}}{
	\sOOhalt{observe(v)}}{}

	\end{array}\]

@@

%	Evaluation \fbox{$ \judgeSeval{C}{e}{o} $}
%	\[
%	\inference{
%		\judgeSreduceTrans{C}{
%			\sOOinspect{e}{\emptyset}{cont(\hckOOmt)}
%		}{
%			\sOOhalt{o}
%		}		
%	}{
%		\judgeSeval{C}{e}{o}
%	}
%	\]

@@

We list a fraction of reduction rules due to space limitation.
When values are constructed, their hypercoercion parts are filled with outputs 
of $ id $. For instance, when a function value is constructed, its first part 
and last part are initialized to $ id(T_1) $ and $ id(T_2) $ respectively.
When evaluating a cast expression, the current continuation is extended with a 
hypercoercion constructed by $ cast $. $ ext $ composes the new hypercoercion 
with the hypercoercion on the top of the continuation.
When evaluating a compound expression, the machine firstly construct the new 
pre-continuation, then turn it to a continuation by adding an identity 
hypercoercion at the top. For instance, when evaluating a \texttt{cons} 
expression, the machine firstly construct $ \hckOOconsI{e_2}{\rho}{\kappa} $, 
the new pre-continuation, then call $ cont $, which adds an identity cast to 
form a continuation. 
When a function call happens, the machine firstly cast the operand. If the 
casting succeeds, the machine then evaluate the body in the extended 
environment and the extended continuation. If the casting fails, the machine 
then halts with the blame label.
